---
title: "Predicting Raw Violent COMPAS scores"
author: "Beau Coker"
date: "6/27/2018"
output: html_document
---

```{r, message=FALSE,warning=FALSE}
library(xgboost)
library(randomForest)
library(tidyverse)
library(lubridate)

source('functions.r')
```

```{r}
load("Table_construction.Rdata")
```

# Violent Recidivism

## Fit age polynomial

```{r}
 age_subset = features %>%
  filter(`Risk of Recidivism_decile_score` != -1, `Risk of Violence_decile_score` != -1) %>% # Filter 1
  select(p_current_age, p_recid_raw = `Risk of Violence_raw_score`) %>%
  filter(p_current_age > 18 & p_current_age <= 70) %>% # Filter 5
  group_by(p_current_age) %>%
  arrange(p_recid_raw, .by_group=TRUE) %>%
  top_n(n=-1, wt=p_recid_raw) # Filter 6

mdl_age = lm(p_recid_raw ~ 
               I(p_current_age^4) + 
               I(p_current_age^3) + 
               I(p_current_age^2) + 
               p_current_age, 
             data=age_subset)


summary(mdl_age)
print("Coefficients:")
sprintf("%.20e",mdl_age$coefficients) # More precision for paper
```

## Generic stuff (applies to all models)

```{r}
### Add useful columns to features and apply row filters used for all models
features_filt = features %>%
  inner_join(
    data_before %>% 
      select(person_id, screening_date, people) %>%
      unnest() %>%
      select(person_id, screening_date, race, sex, name),
    by = c("person_id","screening_date")
  ) %>%
  inner_join(features_on, by = c("person_id","screening_date")) %>%
  inner_join(outcomes, by = c("person_id","screening_date")) %>%
  filter(`Risk of Recidivism_decile_score` != -1, `Risk of Violence_decile_score` != -1) %>% # Filter 1
  filter(!is.na(current_offense_date)) %>% # Filter 3
  mutate(p_recid_raw = `Risk of Violence_raw_score`,
         age_poly = predict(mdl_age, newdata=data.frame(p_current_age=p_current_age)),
         p_recid_raw_noage = p_recid_raw - age_poly)

## Set parameters (each combination will be run)
param <- list(objective = "reg:linear",
              eval_metric = "rmse",
              eta = c(.05,.1),
              gamma = c(.5, 1), 
              max_depth = c(2,5),
              min_child_weight = c(5,10),
              subsample = c(1),
              colsample_bytree = c(1)
)

# svm
param_svm = list(
  type = 'eps-regression',
  cost = c(0.5,1,2),
  epsilon = c(0.5,1,1.5),
  gamma_scale = c(0.5,1,2)
)
```

```{r}
res_rmse = data.frame(Group = 1:4, lm = NA, xgb = NA, rf = NA, svm = NA)
```

```{r}
## Age polynomial
xmin = 18
xmax = 70
xx = seq(xmin,xmax, length.out=1000)

ggplot()+
  geom_point(aes(x=p_current_age, p_recid_raw, color="b"), alpha=.3, data=features_filt) +
  geom_line(aes(x=xx, predict(mdl_age, newdata=data.frame(p_current_age=xx)), color="a")) +
  theme_bw()+
  xlim(xmin,xmax)+
  xlab("Age at COMPAS screening date") +
  ylab("COMPAS violence raw") +
  theme(text = element_text(size=12),
        axis.text=element_text(size=12),
        legend.position="none")
```

```{r}
ggsave("Figures/age_agePoly_violent.pdf",width = 3.5, height = 2.5, units = "in")
```

```{r}
### Number of priors vs. COMPAS remainder
ggplot(data=features_filt) +
  geom_point(aes(x=p_charge_violent, y=p_recid_raw_noage), alpha=.3)+
  theme_bw() +
  xlab("Number of prior violent charges") +
  ylab("COMPAS violence raw remainder") +
  theme(
        text = element_text(size=12),
        axis.text=element_text(size=12))
```

```{r}
ggsave("Figures/priors_rawScoreRemain_violent.pdf",width = 3.5, height = 3.5, units = "in")
```

```{r}
## Total violence history items
features_filt %>%
  mutate(
    p_juv_fel_count = pmin(p_juv_fel_count,2),
    p_felprop_violarrest = pmin(p_felprop_violarrest,5),
    p_murder_arrest = pmin(p_murder_arrest,3),
    p_felassault_arrest = pmin(p_felassault_arrest,3),
    p_misdemassault_arrest = pmin(p_misdemassault_arrest,3),
    #p_famviol_arrest = pmin(p_famviol_arrest,3),
    p_sex_arrest = pmin(p_sex_arrest,3),
    p_weapons_arrest = pmin(p_weapons_arrest,3)) %>%
  ggplot() +
  geom_point(aes(y = p_recid_raw_noage,
                 x=p_juv_fel_count+
                   p_felprop_violarrest+
                   p_murder_arrest+
                   p_felassault_arrest+
                   p_misdemassault_arrest+
                   #p_famviol_arrest+
                   p_sex_arrest+
                   p_weapons_arrest),
             alpha=.3)  +
  xlab("Total Violence History items") +
  ylab("COMPAS violence raw remainder")+
  xlim(0,15) +
  theme_bw() +
  theme(text = element_text(size=12),
        axis.text=element_text(size=12))
```

```{r}
ggsave("Figures/violenceHist_rawScoreRemain_violent.pdf",width = 3.5, height = 3.5, units = "in")
```

## Replicating ProPublica logistic regression

```{r}
propub = features_filt %>%
  filter(current_offense_date <= current_offense_date_limit) %>% # Only people with valid recidivism values
  mutate(age_low = if_else(p_current_age < 25,1,0), 
         age_high = if_else(p_current_age > 45,1,0),
         female = if_else(sex=="Female",1,0),
         n_priors = p_felony_count_person + p_misdem_count_person,
         compas_high = if_else(`Risk of Violence_decile_score` >= 5, 1, 0), # Medium and High risk scores get +1 label
         race = relevel(factor(race), ref="Caucasian")) # Base level is Caucasian, as in ProPublica analysis
```

```{r}
mdl_glm = glm(compas_high ~
                female +
                age_high +
                age_low +
                as.factor(race) +
                p_charge +
                is_misdem +
                recid_violent,
                family=binomial(link='logit'), data=propub)

summary(mdl_glm)
```





## Group 1 models: predicting (raw - age polynomial) without using age variables or race

```{r}
### Create group 1 training data

## Select features and round count features
train = features_filt %>%
  transmute(
    #p_current_age,
    p_age_first_offense,
    p_juv_fel_count = pmin(p_juv_fel_count,2),
    p_felprop_violarrest = pmin(p_felprop_violarrest,5),
    p_murder_arrest = pmin(p_murder_arrest,3),
    p_felassault_arrest = pmin(p_felassault_arrest,3),
    p_misdemassault_arrest = pmin(p_misdemassault_arrest,3),
    #p_famviol_arrest = pmin(p_famviol_arrest,3),
    p_sex_arrest = pmin(p_sex_arrest,3),
    p_weapons_arrest = pmin(p_weapons_arrest,3),
    p_n_on_probation = pmin(p_n_on_probation,5),
    p_current_on_probation = pmin(p_current_on_probation,5),
    p_prob_revoke = pmin(p_prob_revoke,5),
    p_recid_raw_noage)

## Format for xgboost
train_xgb = xgb.DMatrix(
  "data" = train %>% select(-p_recid_raw_noage) %>% as.matrix(),
  "label" = train %>% select(p_recid_raw_noage) %>% as.matrix()
)
```

### Model 1: Linear model

```{r}
mdl_lm = lm(p_recid_raw_noage ~ ., data=train)
summary(mdl_lm)

res_rmse[res_rmse$Group==1,]$lm = rmse(predict(mdl_lm, newdata=train), train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 2: xgboost

```{r}
set.seed(46)
mdl_xgb = fit_xgboost(train_xgb, param)
```

```{r}
### xgboost plot
pred = predict(mdl_xgb, newdata=train_xgb)
actual = train$p_recid_raw_noage

res_rmse[res_rmse$Group==1,]$xgb = rmse(pred, actual) # ADJUST GROUP

axis_min = min(min(pred),min(actual))
axis_max = max(max(pred),max(actual))

data.frame(xgboost = pred, compas=actual) %>%
  ggplot() +
  geom_point(aes(x=compas,y=xgboost), alpha=.3) +
  geom_abline(slope=1, color="red")+
  xlim(c(axis_min,axis_max)) +
  ylim(c(axis_min,axis_max)) +
  coord_fixed() +
  theme_bw()+
  xlab("COMPAS violence raw remainder") +
  ylab("XGBoost prediction")+
  theme(
        text = element_text(size=14),
        axis.text=element_text(size=14))
```

```{r}
### Variable importance
xgb.plot.importance(importance_matrix = xgb.importance(model = mdl_xgb))
```

### Model 3: random forest

```{r}
set.seed(55656)

mdl_rf = randomForest(
  formula = p_recid_raw_noage ~ .,
  data = train
)

res_rmse[res_rmse$Group==1,]$rf = rmse(mdl_rf$predicted, train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 4: SVM

```{r}
mdl_svm = fit_svm(p_recid_raw_noage ~ ., train, param_svm)

res_rmse[res_rmse$Group==1,]$svm = rmse(mdl_svm$fitted, train$p_recid_raw_noage) # ADJUST GROUP
```


### Cleanup
```{r}
rm(train, train_xgb, mdl_lm, mdl_xgb, mdl_rf)
```






## Group 2 models: predicting (raw - age polynomial) without using age variables but with race

```{r}
### Create group 2 training data

## Select features and round count features
train = features_filt %>%
  transmute(
    #p_current_age,
    p_age_first_offense,
    p_juv_fel_count = pmin(p_juv_fel_count,2),
    p_felprop_violarrest = pmin(p_felprop_violarrest,5),
    p_murder_arrest = pmin(p_murder_arrest,3),
    p_felassault_arrest = pmin(p_felassault_arrest,3),
    p_misdemassault_arrest = pmin(p_misdemassault_arrest,3),
    #p_famviol_arrest = pmin(p_famviol_arrest,3),
    p_sex_arrest = pmin(p_sex_arrest,3),
    p_weapons_arrest = pmin(p_weapons_arrest,3),
    p_n_on_probation = pmin(p_n_on_probation,5),
    p_current_on_probation = pmin(p_current_on_probation,5),
    p_prob_revoke = pmin(p_prob_revoke,5),
    race_black = if_else(race=="African-American",1,0),
    race_white = if_else(race=="Caucasian",1,0),
    race_hispanic = if_else(race=="Hispanic",1,0),
    race_asian = if_else(race=="Asian",1,0),
    race_native = if_else(race=="Native American",1,0), # race == "Other" is the baseline
    p_recid_raw_noage)


## Format for xgboost
train_xgb = xgb.DMatrix(
  "data" = train %>% select(-p_recid_raw_noage) %>% as.matrix(),
  "label" = train %>% select(p_recid_raw_noage) %>% as.matrix()
)
```

### Model 1: Linear model

```{r}
mdl_lm = lm(p_recid_raw_noage ~ ., data=train)
summary(mdl_lm)

res_rmse[res_rmse$Group==2,]$lm = rmse(predict(mdl_lm, newdata=train), train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 2: xgboost

```{r}
set.seed(390)
mdl_xgb = fit_xgboost(train_xgb, param)
```

```{r}
### xgboost plot
pred = predict(mdl_xgb, newdata=train_xgb)
actual = train$p_recid_raw_noage

res_rmse[res_rmse$Group==2,]$xgb = rmse(pred, actual) # ADJUST GROUP

axis_min = min(min(pred),min(actual))
axis_max = max(max(pred),max(actual))

data.frame(xgboost = pred, compas=actual) %>%
  ggplot() +
  geom_point(aes(x=compas,y=xgboost), alpha=.3) +
  geom_abline(slope=1, color="red")+
  xlim(c(axis_min,axis_max)) +
  ylim(c(axis_min,axis_max)) +
  coord_fixed() +
  theme_bw()+
  xlab("COMPAS violence raw remainder") +
  ylab("XGBoost prediction")+
  theme(
        text = element_text(size=14),
        axis.text=element_text(size=14))
```



```{r}
### Variable importance
xgb.plot.importance(importance_matrix = xgb.importance(model = mdl_xgb))
```

### Model 3: random forest

```{r}
set.seed(2728)

mdl_rf = randomForest(
  formula = p_recid_raw_noage ~ .,
  data = train
)

res_rmse[res_rmse$Group==2,]$rf = rmse(mdl_rf$predicted, train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 4: SVM

```{r}
mdl_svm = fit_svm(p_recid_raw_noage ~ ., train, param_svm)

res_rmse[res_rmse$Group==2,]$svm = rmse(mdl_svm$fitted, train$p_recid_raw_noage) # ADJUST GROUP
```

### Cleanup
```{r}
rm(train, train_xgb, mdl_lm, mdl_xgb, mdl_rf)
```








## Group 3 models: predicting (raw - age polynomial) without using race but with age variables

```{r}
### Create group 3 training data

## Select features and round count features
train = features_filt %>%
  transmute(
    p_current_age,
    p_age_first_offense,
    p_juv_fel_count = pmin(p_juv_fel_count,2),
    p_felprop_violarrest = pmin(p_felprop_violarrest,5),
    p_murder_arrest = pmin(p_murder_arrest,3),
    p_felassault_arrest = pmin(p_felassault_arrest,3),
    p_misdemassault_arrest = pmin(p_misdemassault_arrest,3),
    #p_famviol_arrest = pmin(p_famviol_arrest,3),
    p_sex_arrest = pmin(p_sex_arrest,3),
    p_weapons_arrest = pmin(p_weapons_arrest,3),
    p_n_on_probation = pmin(p_n_on_probation,5),
    p_current_on_probation = pmin(p_current_on_probation,5),
    p_prob_revoke = pmin(p_prob_revoke,5),
    p_recid_raw_noage)


## Format for xgboost
train_xgb = xgb.DMatrix(
  "data" = train %>% select(-p_recid_raw_noage) %>% as.matrix(),
  "label" = train %>% select(p_recid_raw_noage) %>% as.matrix()
)
```

### Model 1: Linear model

```{r}
mdl_lm = lm(p_recid_raw_noage ~ ., data=train)
summary(mdl_lm)

res_rmse[res_rmse$Group==3,]$lm = rmse(predict(mdl_lm, newdata=train), train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 2: xgboost

```{r}
set.seed(34)
mdl_xgb = fit_xgboost(train_xgb, param)
```

```{r}
### xgboost plot
pred = predict(mdl_xgb, newdata=train_xgb)
actual = train$p_recid_raw_noage

res_rmse[res_rmse$Group==3,]$xgb = rmse(pred, actual) # ADJUST GROUP

axis_min = min(min(pred),min(actual))
axis_max = max(max(pred),max(actual))

data.frame(xgboost = pred, compas=actual) %>%
  ggplot() +
  geom_point(aes(x=compas,y=xgboost), alpha=.3) +
  geom_abline(slope=1, color="red")+
  xlim(c(axis_min,axis_max)) +
  ylim(c(axis_min,axis_max)) +
  coord_fixed() +
  theme_bw()+
  xlab("COMPAS violence raw remainder") +
  ylab("XGBoost prediction")+
  theme(
        text = element_text(size=14),
        axis.text=element_text(size=14))
```

```{r}
### Variable importance
xgb.plot.importance(importance_matrix = xgb.importance(model = mdl_xgb))
```


### Model 3: random forest

```{r}
set.seed(7872)

mdl_rf = randomForest(
  formula = p_recid_raw_noage ~ .,
  data = train
)

res_rmse[res_rmse$Group==3,]$rf = rmse(mdl_rf$predicted, train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 4: SVM

```{r}
mdl_svm = fit_svm(p_recid_raw_noage ~ ., train, param_svm)

res_rmse[res_rmse$Group==3,]$svm = rmse(mdl_svm$fitted, train$p_recid_raw_noage) # ADJUST GROUP
```

### Cleanup
```{r}
rm(train, train_xgb, mdl_lm, mdl_xgb, mdl_rf)
```














## Group 4 models: predicting (raw - age polynomial) using age variables and race

```{r}
### Create group 4 training data

## Select features and round count features
train = features_filt %>%
  transmute(
    p_current_age,
    p_age_first_offense,
    p_juv_fel_count = pmin(p_juv_fel_count,2),
    p_felprop_violarrest = pmin(p_felprop_violarrest,5),
    p_murder_arrest = pmin(p_murder_arrest,3),
    p_felassault_arrest = pmin(p_felassault_arrest,3),
    p_misdemassault_arrest = pmin(p_misdemassault_arrest,3),
    #p_famviol_arrest = pmin(p_famviol_arrest,3),
    p_sex_arrest = pmin(p_sex_arrest,3),
    p_weapons_arrest = pmin(p_weapons_arrest,3),
    p_n_on_probation = pmin(p_n_on_probation,5),
    p_current_on_probation = pmin(p_current_on_probation,5),
    p_prob_revoke = pmin(p_prob_revoke,5),
    race_black = if_else(race=="African-American",1,0),
    race_white = if_else(race=="Caucasian",1,0),
    race_hispanic = if_else(race=="Hispanic",1,0),
    race_asian = if_else(race=="Asian",1,0),
    race_native = if_else(race=="Native American",1,0), # race == "Other" is the baseline
    p_recid_raw_noage)

## Format for xgboost
train_xgb = xgb.DMatrix(
  "data" = train %>% select(-p_recid_raw_noage) %>% as.matrix(),
  "label" = train %>% select(p_recid_raw_noage) %>% as.matrix()
)
```

### Model 1: Linear model

```{r}
mdl_lm = lm(p_recid_raw_noage ~ ., data=train)
summary(mdl_lm)

res_rmse[res_rmse$Group==4,]$lm = rmse(predict(mdl_lm, newdata=train), train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 2: xgboost

```{r}
set.seed(11)
mdl_xgb = fit_xgboost(train_xgb, param)
```

```{r}
### xgboost plot
pred = predict(mdl_xgb, newdata=train_xgb)
actual = train$p_recid_raw_noage

res_rmse[res_rmse$Group==4,]$xgb = rmse(pred, actual) # ADJUST GROUP

axis_min = min(min(pred),min(actual))
axis_max = max(max(pred),max(actual))

data.frame(xgboost = pred, compas=actual) %>%
  ggplot() +
  geom_point(aes(x=compas,y=xgboost), alpha=.3) +
  geom_abline(slope=1, color="red")+
  xlim(c(axis_min,axis_max)) +
  ylim(c(axis_min,axis_max)) +
  coord_fixed() +
  theme_bw()+
  xlab("COMPAS violence raw remainder") +
  ylab("Prediction of COMPAS violence raw remainder")+
  theme(
        text = element_text(size=12),
        axis.text=element_text(size=12))
```

```{r}
ggsave("Figures/rawScoreRemain_xgboost_violent.pdf",width = 4, height = 4, units = "in")
```

```{r}
### Variable importance
xgb.plot.importance(importance_matrix = xgb.importance(model = mdl_xgb))
```

```{r}
highlight = data.frame(
  person_id= c(799, 1284, 1394, 1497, 1515, 1638, 3145, 3291, 5722, 6337, 6886, 7997, 8200, 8375, 8491, 10553, 10774, 11231, 11312, 11414),
  screening_date = ymd(c("2014-06-15","2014-05-14","2014-11-28","2013-07-29","2013-10-23","2013-10-04","2014-12-14","2013-01-17","2013-10-24","2014-02-04","2013-07-12","2014-04-26","2014-05-05","2013-03-19","2014-01-18","2014-09-20","2013-04-09","2014-02-23","2014-05-02","2014-11-26")),
  highlight = TRUE
)

df_plot = features_filt %>%
  bind_cols(xgboost = predict(mdl_xgb, newdata=train_xgb)) %>%
  left_join(highlight, by = c("person_id","screening_date")) %>%
  mutate(highlight = if_else(is.na(highlight), FALSE, TRUE)) %>%
  mutate(highlight = factor(if_else(highlight==TRUE,"In Table 5", "Not in Table 5"), levels=c("In Table 5", "Not in Table 5")))

person_id_text_topright = c(8491, 8375, 1497)
#person_id_text_topright = highlight$person_id
person_id_text_topleft = c(5722, 11231)
person_id_text_botright = c(799, 11312, 1284, 11414)
person_id_text_botleft = c()

ggplot() +
  geom_point(aes(x=xgboost,y=p_recid_raw, color=highlight),  alpha = .3, data = filter(df_plot, highlight=="Not in Table 5")) +
  geom_point(aes(x=xgboost,y=p_recid_raw, color=highlight),  data = filter(df_plot, highlight=="In Table 5")) +
  theme_bw()+
  geom_text(aes(x=xgboost,y=p_recid_raw,label=name),size=3,nudge_x=0, nudge_y=0, hjust="left",vjust="bottom", data=filter(df_plot, person_id %in% person_id_text_topright & highlight=="In Table 5")) + 
  geom_text(aes(x=xgboost,y=p_recid_raw,label=name),size=3,nudge_x=0, nudge_y=0, hjust="right",vjust="bottom", data=filter(df_plot, person_id %in% person_id_text_topleft & highlight=="In Table 5")) + 
  geom_text(aes(x=xgboost,y=p_recid_raw,label=name),size=3,nudge_x=0, nudge_y=0, hjust="left",vjust="top", data=filter(df_plot, person_id %in% person_id_text_botright & highlight=="In Table 5")) + 
  geom_text(aes(x=xgboost,y=p_recid_raw,label=name),size=3,nudge_x=0, nudge_y=0, hjust="right",vjust="top", data=filter(df_plot, person_id %in% person_id_text_botleft & highlight=="In Table 5")) + 
  xlab("Prediction of COMPAS violence raw remainder") +
  ylab("COMPAS violence raw")+
  theme(
    text = element_text(size=12),
    axis.text=element_text(size=12),
    #legend.position = "top",
    legend.position="none") +
  scale_color_discrete(name = element_blank())
```


```{r}
ggsave("Figures/xgboost_rawScore_violent.pdf",width = 4, height = 4, units = "in")
```

### Model 3: random forest

```{r}
set.seed(379)

mdl_rf = randomForest(
  formula = p_recid_raw_noage ~ .,
  data = train
)

res_rmse[res_rmse$Group==4,]$rf = rmse(mdl_rf$predicted, train$p_recid_raw_noage) # ADJUST GROUP
```

### Model 4: SVM

```{r}
mdl_svm = fit_svm(p_recid_raw_noage ~ ., train, param_svm)

res_rmse[res_rmse$Group==4,]$svm = rmse(mdl_svm$fitted, train$p_recid_raw_noage) # ADJUST GROUP
```

### Cleanup
```{r}
rm(train, train_xgb, mdl_lm, mdl_xgb, mdl_rf)
```










## Comparison

```{r}
knitr::kable(res_rmse)
```




